Expert pianists make specific exaggerations for teaching

### 2022-12-12 paper/manuscript.Rmd
finalised manuscript (published)
Signed-off-by: Atsuko Tominaga <Tominaga_Atsuko@phd.ceu.edu>

Deliberate teaching has supported human skill transmission over generations and provides a key route for learning [@thornton_2008; @tomasello_2016]. Experts modulate their behaviour so that novices can extract relevant information to learn a novel skill. For instance, adults modulate their speech (motherese) and actions (motionese) when demonstrating a skill to infant learners [@brand_2002; @saint-georges_2013]. These modulations include slowing down and exaggerating sounds and actions. Similar findings were obtained in studies with adult learners, where exaggerations were observed when native British English speakers were talking to first language English learners [@uther_2007] and when skilled adults were teaching xylophone melodies to novices [@mcellin_2017]. The observed modulations are thought to provide communicative signals that can facilitate learning by affecting attention and memory [@csibra_2009].

@@

We recruited 36 piano experts who played the piano for at least the past 10 years or were studying advanced piano performance at a music school at the time of recruitment. For data analysis, we excluded three participants due to experimental errors, and two participants because they deviated substantially from the prescribed tempo (outside 2 standard deviations from the average tempo across participants). Thirty-one participants (`r nrow(dt_included[Gender == "Female"])` female) were included in data analysis. Most participants were right-handed (left: `r nrow(dt_included[Handedness == "Left"])`, ambidextrous: `r nrow(dt_included[Handedness == "Both"])`) with a mean age of `r round(mean(dt_included$Age), 2)` (*SD* = `r round(sd(dt_included$Age), 2)`). They had `r round(mean(dt_included$PianoTotalPractice), 2)` years of practice on average (*SD* = `r round(sd(dt_included$PianoTotalPractice), 2)`). `r nrow(dt_included[TeachingPiano == "Yes"])` participants had an experience in teaching the piano (*M* = `r round( mean(dt_included[TeachingPiano == "Yes"]$TeachingPianoYears), 2)` years, *SD* = `r round( sd(dt_included[TeachingPiano == "Yes"]$TeachingPianoYears), 2)`). All participants gave their informed consent before the experiment started and received vouchers for their participation. The study (No. 2018-124) was approved by the United Ethical Review Committee for Research in Psychology (EPKEB) in Hungary.

@@

One musical excerpt was used as a stimulus. The excerpt was taken from ``A Dozen a Day - Play with Ease in Many Keys" by Edna-Mae Burnam and modified for the experiment. It consisted of a 6-measure isochronous melody noted in 4/4 meter. The stimulus was composed in C major to be played with the right hand only. Original sheet music (i.e., sheet music without expressive notations, *Fig \ref{fig:stimuli} A*) was used for the purpose of practice. Expressive notations were added to the original sheet music for the experiment. They referred to either articulation or dynamics (*Fig \ref{fig:stimuli} B-C*). Articulation was notated as either legato or staccato. Legato indicates that musical notes are to be connected and should sound smooth. Staccato requires producing musical notes with shortened duration, keeping them separate from each other. Dynamics was notated as either forte or piano. Forte indicates that musical notes should be played loudly whereas piano indicates that musical notes should be played softly. The notation did not include any indication of fingering (i.e., the positioning of the fingers when playing the piano) because the piece was simple and pilot testing had shown that specifying fingering was not necessary.

@@

To compare the mean IOIs between the teaching and performing condition, we conducted a Wilcoxon Signed-rank test, instead of a paired *t*-test, because a Shapiro-Wilk test showed that the distribution of the mean difference was significantly different from the normal distribution (*p* `r if(ioi_dyn_norm$p.value < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', ioi_dyn_norm$p.value))}`). The Wilcoxon Signed-rank test revealed no significant difference between the teaching condition [*Mdn* = `r ioi_dyn_all[Condition == "teaching"]$Median` (ms), *IQR* = `r ioi_dyn_all[Condition == "teaching"]$IQR`] and the performing condition [*Mdn* = `r ioi_dyn_all[Condition == "performing"]$Median` (ms), *IQR* = `r ioi_dyn_all[Condition == "performing"]$IQR`] while playing the piece with the notated dynamics (*p* `r if(ioi_dyn_wilcoxon$p.value < 0.001){"< 0.001"} else {paste("= ", sprintf('%.2f', ioi_dyn_wilcoxon$p.value))}`, *r* = `r round(wilcoxeff_dyn$effsize, 2)`, two-tailed, *Fig \ref{fig:ioi-1}*).

@@

A two-way repeated-measures ANOVA with the factors Condition (teaching vs. performing) and Dynamics (forte vs. piano) showed that there was a significant main effect of Dynamics (*F*(`r vel_dyn_aov$anova_table[2,1]`,`r vel_dyn_aov$anova_table[2,2]`) = `r sprintf('%.0f', vel_dyn_aov$anova_table[2,4])`, *p* `r if(vel_dyn_aov$anova_table[2,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_dyn_aov$anova_table[2,6]))}`, $\eta_G^2$ = `r sprintf('%.2f', vel_dyn_aov$anova_table[2,5])`) and a significant interaction between Condition and Dynamics (*F*(`r vel_dyn_aov$anova_table[3,1]`,`r vel_dyn_aov$anova_table[3,2]`) = `r sprintf('%.2f', vel_dyn_aov$anova_table[3,4])`, *p* `r if(vel_dyn_aov$anova_table[3,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_dyn_aov$anova_table[3,6]))}`, $\eta_G^2$ = `r sprintf('%.3f', vel_dyn_aov$anova_table[3,5])`, *Fig \ref{fig:vel-1}*). However, post-hoc comparisons based on the estimated marginal means with Tukey adjustment did not find a significant difference between the teaching [Forte: *M* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Forte"]$Mean`, *SD* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Forte"]$SD`, Piano: *M* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Piano"]$Mean`, *SD* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Piano"]$SD`] and the performing condition [Forte: *M* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Forte"]$Mean`, *SD* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Forte"]$SD`, Piano: *M* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Piano"]$Mean`, *SD* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Piano"]$SD`] for each subcomponent (Forte: *p* `r if(vel_dyn_posthoc$p.value[1] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_dyn_posthoc$p.value[1]))}`, Piano: *p* `r if(vel_dyn_posthoc$p.value[2] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.2f', vel_dyn_posthoc$p.value[2]))}`).

@@

The results for KOT and KV showed that participants successfully highlighted relevant aspects of articulation and dynamics. Specifically, participants exaggerated staccato when teaching articulation and forte when teaching dynamics. Moreover, our exploratory analysis demonstrated that participants made a larger contrast in dynamics at transition points (i.e., forte to piano or piano to forte). Importantly, participants did not modulate their performance in terms of irrelevant aspects of the techniques for teaching purposes (e.g., modulating the smoothness of sound while teaching dynamics). These findings confirmed that participants modulated their performance in systematic and fine-grained ways to teach particular techniques.

@@

We recruited 21 piano experts who already had a degree (above bachelor or equivalent) in piano performance/teaching or were studying advanced piano performance at a music school at the time of recruitment. For data analysis, we excluded one participant due to insufficient motor skills. Twenty participants (`r nrow(dt_2_included[Gender == "Female"])` female) were included in the data analysis. Most participants were right-handed (left: `r nrow(dt_2_included[Handedness == "Left"])`) with a mean age of `r round(mean(dt_2_included$Age), 2)` (*SD* = `r round(sd(dt_2_included$Age), 2)`). They had `r round(mean(dt_2_included$PianoTotalPractice), 2)` years of practice on average (*SD* = `r round(sd(dt_2_included$PianoTotalPractice), 2)`). `r nrow(dt_2_included[TeachingPiano == "Yes"])` participants had an experience in teaching the piano (*M* = `r round( mean(dt_2_included[TeachingPiano == "Yes"]$TeachingPianoYears), 2)` years, *SD* = `r round( sd(dt_2_included[TeachingPiano == "Yes"]$TeachingPianoYears), 2)`). All participants gave their informed consent before the experiment started and received vouchers for their participation. The study (No. 2018-124) was approved by the United Ethical Review Committee for Research in Psychology (EPKEB) in Hungary.

@@

The same apparatus as in Experiment 1 was used. We selected Clementi's Sonatina Op.36 (No.3) in C major as a stimulus because it contains our targeted expressions (i.e., articulation and dynamics) and is relatively simple in terms of motor skills. The first 12 measures of the original piece were used and modified so that the piece had an almost equal number of data points for each dependent variable. The modified piece consisted of a 12-measure isochronous melody notated in 4/4 meter to be played with the right hand only. Original sheet music was used for the purpose of practice (*Fig \ref{fig:stimuli-2} A*). Expressive notations were added to the original sheet music for the experiment (*Fig \ref{fig:stimuli-2} B-C*). These excerpts were confirmed to be musically natural by a doctoral student in piano performance at Liszt Ferenc Academy of Music in Hungary. The fingering was also assigned and confirmed by the same doctoral student.

@@

We employed the same procedure as in Experiment 1 with several modifications. First, all participants were required to memorise the piece without expressive notations (i.e., *Fig \ref{fig:stimuli-2} A*) prior to the experiment so that they had enough time to familiarise themselves with the piece and perform it without pitch errors while performing it with different notated expressions in the lab. Second, we modified the wording of instructions for the performing condition so that both instructions had the same focus on expressive notations because there had been less emphasis on expression in the instruction in the performing condition in Experiment 1 (see details in *[Supplementary Material](#supplementary)* 1). Third, participants could choose their preferred tempo from one of three options (100, 110 and 120 quarter beats per minute) because some evidence show that musicians have their own preferred tempo, which affects their temporal variability [@zamm_2018]. The chosen tempo was again cued by a leading metronome.  To make sure that participants memorised the piece and had sufficient motor skills, we asked participants to perform the piece without looking at the original sheet music (*Fig \ref{fig:stimuli-2} A*). Those who could not perform the piece without pitch errors twice consecutively within five attempts were excluded from data analysis (as a result, one participant was excluded). The rest of the procedure was identical to Experiment 1.

@@

The results from Experiment 2 replicate our earlier findings and provide further evidence that skilled pianists modulated their performance in specific ways for teaching purposes. As in Experiment 1, we observed slower performance only while teaching the notated articulation. Again, KORs showed that when teaching articulation, participants exaggerated staccato while there was no significant difference in legato between the two conditions. While teaching dynamics, KV showed that participants exaggerated forte and piano. This pattern of exaggeration was more pronounced than in Experiment 1, where only higher KV for forte was observed. Furthermore, we again found that when teaching dynamics, at transition points participants produced a larger contrast in dynamics between forte and piano, bidirectionally. Importantly, participants did not modulate performance aspects that were irrelevant for the techniques to be taught (i.e., no modulation of dynamics when teaching articulation and vice versa). Overall, Experiment 2 demonstrated systematic and fine-grained didactic performance modulations for a naturalistic piece of music.

@@

We quantified participants’ sound modulations in terms of tempo, articulation and dynamics so as to examine if and how participants modulate their sound. However, it is possible that participants show didactic modulations in other aspects such as temporal variability and temporal grouping (see *[Supplementary Material](#supplementary)* 3 and 4). For example, participants tended to slow down at transition points regardless of the techniques they were teaching although overall, we found that participants played slower only when teaching the piece with the notated articulation. Future research is needed to investigate how musicians make particular modulations in relation to musical structures.

Musicians and music teachers tend to consider expressive skills as performers’ most important skills [@lindstrom_2003]. In the present study, experts focused on the teaching of expressions that were notated. However, expression also has other facets [@juslin_2003] that are not only piece-related (e.g., notated expressions) but performer- or context-related (e.g., performer’s interpretation, listening contexts such as recordings or concerts). The ultimate goal of expressive performance is considered to lie in developing performers’ own interpretations of music and convey affects and emotions by using expressive tools such as articulation, dynamics, tempo and timing [@meissner_2021]. The current research opens up the possibility of studying the teaching of expressive performance by demonstrating that pianists could signal didactic intentions without relying on verbal instructions, which are heavily used in actually teaching settings [@woody_2000; @colprit_2000]. Given that students often learn by listening to recordings [@volioti_2017], it is important to investigate how performance itself could affect students’ learning processes.It would be interesting to examine how teachers play (i.e., demonstrate) relates to verbal instructions they give.

The present findings extend earlier research on teaching-related action modulations. First, our study sheds light on the teaching of technical skills that are an integral part of skill acquisition in artistic domains. We showed that compared to an expressive performance baseline, experts made specific modulations to teach particular techniques. Second, the specificity of the observed modulations supports the idea that teaching comprises more than generic modulations like slowing down or overall exaggeration that may draw learners’ attention. Rather, expert demonstrators seem to follow principles of relevance in communication [@sperber_1995], highlighting only those aspects they are intending to demonstrate. How learners benefit from the perceptual and motor cues that come with specific exaggerations, and whether understanding the teacher's intentions explicitly adds to the learning success are important questions for future research.

@@

apa_table(
  ioi_all[, -c(8, 9)]
  , caption = "Experiment 1: Comparison between those with and without piano teaching experience in terms of IOIs (ms)."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.1 B) or dynamics (Fig.1 C). TeachingExp indicates if participants had an experience in teaching the piano."
  , escape = TRUE
)
```

@@

apa_table(
  kot_art_all
  , caption = "Experiment 1: Comparison between those with and without piano teaching experience in terms of KOT (ms)."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.1 B) or dynamics (Fig.1 C). TeachingExp indicates if participants had an experience in teaching the piano."
  , escape = TRUE
)
```

@@

apa_table(
  vel_dyn_all
  , caption = "Experiment 1: Comparison between those with and without piano teaching experience in terms of KV (0-127)."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.1 B) or dynamics (Fig.1 C). TeachingExp indicates if participants had an experience in teaching the piano."
  , escape = TRUE
)
```

@@

apa_table(
  vel_diff_dyn_all
  , caption = "Experiment 1: Comparison between those with and without piano teaching experience in terms of KV Difference at transition points."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.1 B) or dynamics (Fig.1 C). TeachingExp indicates if participants had an experience in teaching the piano."
  , escape = TRUE
)
```

@@

apa_table(
  ioi_all[, -c(8, 9)]
  , caption = "Experiment 2: Comparison between those with and without piano teaching experience in terms of normalised IOIs."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.5 B) or dynamics (Fig.5 C). TeachingExp indicates if participants had an experience in teaching the piano."
  , escape = TRUE
  , digits = 3
)
```

@@

apa_table(
  kor_art_all
  , caption = "Experiment 2: Comparison between those with and without piano teaching experience in terms of KOR."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.5 B) or dynamics (Fig.5 C). TeachingExp indicates if participants had an experience in teaching the piano."
  , escape = TRUE
  , digits = 3
)
```

@@

apa_table(
  vel_dyn_all
  , caption = "Experiment 2: Comparison between those with and without piano teaching experience in terms of KV (0-127)."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.5 B) or dynamics (Fig.5 C). TeachingExp indicates if participants had an experience in teaching the piano."
  , escape = TRUE
)
```

@@

apa_table(
  vel_diff_dyn_all
  , caption = "Experiment 2: Comparison between those with and without piano teaching experience in terms of KV Difference at transition points."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.5 B) or dynamics (Fig.5 C). TeachingExp indicates if participants had an experience in teaching the piano."
  , escape = TRUE
)
```

>>>>

Deliberate teaching has supported human skill transmission over generations and provides a key route for learning [@thornton_2008; @tomasello_2016]. Experts modulate their behaviour so that novices can extract relevant information to learn a novel skill. For instance, adults modulate their speech (motherese) and actions (motionese) when demonstrating a skill to infant learners [@brand_2002; @saint-georges_2013]. These modulations include slowing down and exaggerating sounds and actions. Similar findings were obtained in studies with adult learners, where exaggerations were observed when native British English speakers were talking to second language English learners [@uther_2007] and when skilled adults were teaching xylophone melodies to novices [@mcellin_2017]. The observed modulations are thought to provide communicative signals that can facilitate learning by affecting attention and memory [@csibra_2009].

@@

We recruited 36 piano experts who played the piano for at least the past 10 years or were studying advanced piano performance at a music school at the time of recruitment. For data analysis, we excluded three participants due to experimental errors, and two participants because they deviated substantially from the prescribed tempo (outside 2 standard deviations from the average tempo across participants). Thirty-one participants (`r nrow(dt_included[Gender == "Female"])` female) were included in data analysis. Most participants were right-handed (left: `r nrow(dt_included[Handedness == "Left"])`, ambidextrous: `r nrow(dt_included[Handedness == "Both"])`) with a mean age of `r round(mean(dt_included$Age), 2)` (*SD* = `r round(sd(dt_included$Age), 2)`). They had `r round(mean(dt_included$PianoTotalPractice), 2)` years of practice on average (*SD* = `r round(sd(dt_included$PianoTotalPractice), 2)`). `r nrow(dt_included[TeachingPiano == "Yes"])` participants had experience in teaching the piano (*M* = `r round( mean(dt_included[TeachingPiano == "Yes"]$TeachingPianoYears), 2)` years, *SD* = `r round( sd(dt_included[TeachingPiano == "Yes"]$TeachingPianoYears), 2)`). All participants gave their informed consent before the experiment started and received vouchers for their participation. The study (No. 2018-124) was approved by the United Ethical Review Committee for Research in Psychology (EPKEB) in Hungary.

@@

One musical excerpt was used as a stimulus. The excerpt was taken from ``A Dozen a Day - Play with Ease in Many Keys" by Edna-Mae Burnam and modified for the experiment. It consisted of a 6-measure isochronous melody noted in a 4/4 metre. The stimulus was composed in C major to be played with the right hand only. Original sheet music (i.e., sheet music without expressive notations, *Fig \ref{fig:stimuli} A*) was used for the purpose of practice. Expressive notations were added to the original sheet music for the experiment. They referred to either articulation or dynamics (*Fig \ref{fig:stimuli} B-C*). Articulation was notated as either legato or staccato. Legato indicates that musical notes are to be connected and should sound smooth. Staccato requires producing musical notes with shortened duration, keeping them separate from each other. Dynamics was notated as either forte or piano. Forte indicates that musical notes should be played loudly whereas piano indicates that musical notes should be played softly. The notation did not include any indication of fingering (i.e., the positioning of the fingers when playing the piano) because the piece was simple and pilot testing had shown that specifying fingering was not necessary.

@@

To compare the mean IOIs between the teaching and performing condition, we conducted a Wilcoxon Signed-rank test, instead of a paired *t*-test, because a Shapiro-Wilk test showed that the distribution of the mean difference was significantly different from the normal distribution (*p* `r if(ioi_dyn_norm$p.value < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', ioi_dyn_norm$p.value))}`). The Wilcoxon Signed-rank test revealed no significant difference between the teaching condition [*Mdn* = `r ioi_dyn_all[Condition == "teaching"]$Median` (ms), *IQR* = `r ioi_dyn_all[Condition == "teaching"]$IQR`] and performing condition [*Mdn* = `r ioi_dyn_all[Condition == "performing"]$Median` (ms), *IQR* = `r ioi_dyn_all[Condition == "performing"]$IQR`] while playing the piece with the notated dynamics (*p* `r if(ioi_dyn_wilcoxon$p.value < 0.001){"< 0.001"} else {paste("= ", sprintf('%.2f', ioi_dyn_wilcoxon$p.value))}`, *r* = `r round(wilcoxeff_dyn$effsize, 2)`, two-tailed, *Fig \ref{fig:ioi-1}*).

@@

A two-way repeated-measures ANOVA with the factors Condition (teaching vs. performing) and Dynamics (forte vs. piano) showed that there was a significant main effect of Dynamics (*F*(`r vel_dyn_aov$anova_table[2,1]`,`r vel_dyn_aov$anova_table[2,2]`) = `r sprintf('%.0f', vel_dyn_aov$anova_table[2,4])`, *p* `r if(vel_dyn_aov$anova_table[2,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_dyn_aov$anova_table[2,6]))}`, $\eta_G^2$ = `r sprintf('%.2f', vel_dyn_aov$anova_table[2,5])`) and a significant interaction between Condition and Dynamics (*F*(`r vel_dyn_aov$anova_table[3,1]`,`r vel_dyn_aov$anova_table[3,2]`) = `r sprintf('%.2f', vel_dyn_aov$anova_table[3,4])`, *p* `r if(vel_dyn_aov$anova_table[3,6] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_dyn_aov$anova_table[3,6]))}`, $\eta_G^2$ = `r sprintf('%.3f', vel_dyn_aov$anova_table[3,5])`, *Fig \ref{fig:vel-1}*). However, post-hoc comparisons based on the estimated marginal means with Tukey adjustment did not find a significant difference between the teaching [Forte: *M* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Forte"]$Mean`, *SD* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Forte"]$SD`, Piano: *M* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Piano"]$Mean`, *SD* = `r vel_dyn_all[Condition == "teaching" & Subcomponent == "Piano"]$SD`] and performing condition [Forte: *M* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Forte"]$Mean`, *SD* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Forte"]$SD`, Piano: *M* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Piano"]$Mean`, *SD* = `r vel_dyn_all[Condition == "performing" & Subcomponent == "Piano"]$SD`] for each subcomponent (Forte: *p* `r if(vel_dyn_posthoc$p.value[1] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.3f', vel_dyn_posthoc$p.value[1]))}`, Piano: *p* `r if(vel_dyn_posthoc$p.value[2] < 0.001){"< 0.001"} else {paste("= ", sprintf('%.2f', vel_dyn_posthoc$p.value[2]))}`).

@@

The results for KOT and KV showed that participants successfully highlighted relevant aspects of articulation and dynamics. Specifically, participants exaggerated staccato when teaching articulation. Moreover, our exploratory analysis demonstrated that participants made a larger contrast in dynamics at transition points (i.e., forte to piano or piano to forte). Importantly, participants did not modulate their performance in terms of irrelevant aspects of the techniques for teaching purposes (e.g., modulating the smoothness of sound while teaching dynamics). These findings confirmed that participants modulated their performance in systematic and fine-grained ways to teach particular techniques.

@@

We recruited 21 piano experts who already had a degree (above bachelor or equivalent) in piano performance/teaching or were studying advanced piano performance at a music school at the time of recruitment. For data analysis, we excluded one participant due to insufficient motor skills. Twenty participants (`r nrow(dt_2_included[Gender == "Female"])` female) were included in the data analysis. Most participants were right-handed (left: `r nrow(dt_2_included[Handedness == "Left"])`) with a mean age of `r round(mean(dt_2_included$Age), 2)` (*SD* = `r round(sd(dt_2_included$Age), 2)`). They had `r round(mean(dt_2_included$PianoTotalPractice), 2)` years of practice on average (*SD* = `r round(sd(dt_2_included$PianoTotalPractice), 2)`). `r nrow(dt_2_included[TeachingPiano == "Yes"])` participants had experience in teaching the piano (*M* = `r round( mean(dt_2_included[TeachingPiano == "Yes"]$TeachingPianoYears), 2)` years, *SD* = `r round( sd(dt_2_included[TeachingPiano == "Yes"]$TeachingPianoYears), 2)`). All participants gave their informed consent before the experiment started and received vouchers for their participation. The study (No. 2018-124) was approved by the United Ethical Review Committee for Research in Psychology (EPKEB) in Hungary.

@@

The same apparatus as in Experiment 1 was used. We selected Clementi's Sonatina Op.36 (No.3) in C major as a stimulus because it contains our targeted expressions (i.e., articulation and dynamics) and is relatively simple in terms of motor skills. The first 12 measures of the original piece were used and modified so that the piece had an almost equal number of data points for each dependent variable. The modified piece consisted of a 12-measure isochronous melody notated in a 4/4 metre to be played with the right hand only. Original sheet music was used for the purpose of practice (*Fig \ref{fig:stimuli-2} A*). Expressive notations were added to the original sheet music for the experiment (*Fig \ref{fig:stimuli-2} B-C*). These excerpts were confirmed to be musically natural by a doctoral student in piano performance at Liszt Ferenc Academy of Music in Hungary. The fingering was also assigned and confirmed by the same doctoral student.

@@

We employed the same procedure as in Experiment 1 with several modifications. First, all participants were required to memorise the piece without expressive notations (i.e., *Fig \ref{fig:stimuli-2} A*) prior to the experiment so that they had enough time to familiarise themselves with the piece and perform it without pitch errors while performing it with different notated expressions in the lab. Second, we modified the wording of instructions for the performing condition so that both instructions had the same focus on expressive notations because there had been less emphasis on expression in the instruction in the performing condition in Experiment 1 (see details in *[Supplementary Material](#supplementary)* 1). Third, participants could choose their preferred tempo from one of three options (100, 110 and 120 quarter beats per minute) because some evidence shows that musicians have their own preferred tempo, which affects their temporal variability [@zamm_2018]. The chosen tempo was again cued by a leading metronome.  To make sure that participants memorised the piece and had sufficient motor skills, we asked participants to perform the piece without looking at the original sheet music (*Fig \ref{fig:stimuli-2} A*). Those who could not perform the piece without pitch errors twice consecutively within five attempts were excluded from data analysis (as a result, one participant was excluded). The rest of the procedure was identical to Experiment 1.

@@

The results from Experiment 2 replicate our earlier findings and provide further evidence that skilled pianists modulated their performance in specific ways for teaching purposes. As in Experiment 1, we observed slower performance only while teaching the notated articulation. Again, KORs showed that when teaching articulation, participants exaggerated staccato while there was no significant difference in legato between the two conditions. While teaching dynamics, KV showed that participants exaggerated forte and piano. This pattern of exaggeration was more pronounced than in Experiment 1, where post-hoc comparions did not find significant differences between the two conditions for each subcomponent. Furthermore, we again found that when teaching dynamics, at transition points participants produced a larger contrast in dynamics between forte and piano, bidirectionally. Importantly, participants did not modulate performance aspects that were irrelevant for the techniques to be taught (i.e., no modulation of dynamics when teaching articulation and vice versa). Overall, Experiment 2 demonstrated systematic and fine-grained didactic performance modulations for a naturalistic piece of music.

@@

We quantified participants’ sound modulations in terms of tempo, articulation and dynamics so as to examine if and how participants modulate their sound. However, it is possible that participants show didactic modulations in other aspects such as temporal variability and temporal grouping (see *[Supplementary Materials](#supplementary)* 3 and 4). For example, participants tended to slow down at transition points regardless of the techniques they were teaching although overall, we found that participants played slower only when teaching the piece with the notated articulation. Future research is needed to investigate how musicians make particular modulations in relation to musical structures.

Musicians and music teachers tend to consider expressive skills as performers’ most important skills [@lindstrom_2003]. In the present study, experts focused on the teaching of specific expressions that were notated. However, expression also has other facets [@juslin_2003] that are not only piece-related (e.g., notated expressions) but performer- or context-related (e.g., performer’s interpretation, listening contexts such as recordings or concerts). The ultimate goal of expressive performance is considered to lie in developing performers’ own interpretations of music and convey affects and emotions by using expressive tools such as articulation, dynamics, tempo and timing [@meissner_2021]. The current research opens up the possibility of studying the teaching of expressive performance by demonstrating that pianists could signal didactic intentions without relying on verbal instructions, which are heavily used in actual teaching settings [@woody_2000; @colprit_2000]. Given that students often learn by listening to recordings [@volioti_2017], it is important to investigate how performance itself could affect students’ learning processes. It would also be interesting to examine how teachers’ playing (i.e., demonstrations) relates to the verbal instructions they give.

The present findings extend earlier research on teaching-related action modulations. First, our study sheds light on the teaching of technical skills that are an integral part of skill acquisition in artistic domains. We showed that compared to an expressive performance baseline, experts made specific modulations to teach particular techniques. Second, the specificity of the observed modulations supports the idea that teaching comprises more than generic modulations like slowing down or overall exaggeration that may draw learners’ attention. Rather, expert demonstrators seem to follow principles of relevance in communication [@sperber_1995], highlighting only those aspects they are intending to demonstrate. How learners benefit from the perceptual and motor cues that come with specific exaggerations, and whether understanding the teacher’s intentions explicitly adds to the learning success are important questions for future research.

@@

apa_table(
  ioi_all[, -c(8, 9)]
  , caption = "Experiment 1: Comparison between those with and without piano teaching experience in terms of IOIs (ms)."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.1 B) or dynamics (Fig.1 C). TeachingExp indicates if participants had experience in teaching the piano."
  , escape = TRUE
)
```

@@

apa_table(
  kot_art_all
  , caption = "Experiment 1: Comparison between those with and without piano teaching experience in terms of KOT (ms)."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.1 B) or dynamics (Fig.1 C). TeachingExp indicates if participants had experience in teaching the piano."
  , escape = TRUE
)
```

@@

apa_table(
  vel_dyn_all
  , caption = "Experiment 1: Comparison between those with and without piano teaching experience in terms of KV (0-127)."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.1 B) or dynamics (Fig.1 C). TeachingExp indicates if participants had experience in teaching the piano."
  , escape = TRUE
)
```

@@

apa_table(
  vel_diff_dyn_all
  , caption = "Experiment 1: Comparison between those with and without piano teaching experience in terms of KV Difference at transition points."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.1 B) or dynamics (Fig.1 C). TeachingExp indicates if participants had experience in teaching the piano."
  , escape = TRUE
)
```

@@

apa_table(
  ioi_all[, -c(8, 9)]
  , caption = "Experiment 2: Comparison between those with and without piano teaching experience in terms of normalised IOIs."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.5 B) or dynamics (Fig.5 C). TeachingExp indicates if participants had experience in teaching the piano."
  , escape = TRUE
  , digits = 3
)
```

@@

apa_table(
  kor_art_all
  , caption = "Experiment 2: Comparison between those with and without piano teaching experience in terms of KOR."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.5 B) or dynamics (Fig.5 C). TeachingExp indicates if participants had experience in teaching the piano."
  , escape = TRUE
  , digits = 3
)
```

@@

apa_table(
  vel_dyn_all
  , caption = "Experiment 2: Comparison between those with and without piano teaching experience in terms of KV (0-127)."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.5 B) or dynamics (Fig.5 C). TeachingExp indicates if participants had experience in teaching the piano."
  , escape = TRUE
)
```

@@

apa_table(
  vel_diff_dyn_all
  , caption = "Experiment 2: Comparison between those with and without piano teaching experience in terms of KV Difference at transition points."
  , note = "N, SD and SEM represent sample size, standard deviation and standard error of the mean, respectively. Technique indicates performance data when participants playing the piece with either articulation (Fig.5 B) or dynamics (Fig.5 C). TeachingExp indicates if participants had experience in teaching the piano."
  , escape = TRUE
)
```
